---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NicheScribe

<!-- badges: start -->
<!-- badges: end -->

**NicheScribe** is a hierarchical classifier for murine bone marrow stromal cells based on 10x single cell RNA-sequencing (scRNA-seq) data. Our approach permits rapid, consistent, accurate, and generalizable annotation of stromal cell types by leveraging curated reference atlases and gene expression signatures, and facilitates comparative analysis of the same cell type across conditions. <!-- For more information and applications of NicheScribe, see our accompanying manuscript... -->

NicheScribe:

* extracts "nonhematopoietic" cells based on an extensive reference atlas spanning whole bone marrow, hematopoietic stem and progenitor cells (HSPCs), and stromal compartments (central marrow and endosteum), 
* classifies stromal cells using either
  1. a reference derived from 10 FACS-purified stromal populations, or
  2. a manually curated reference based on key marker genes, 
* projects stromal cells onto a reference UMAP embedding for easy visualization, and
* scores inflammation state in leptin receptor-positive mesenchymal stromal cells (MSC-L) from differentially upregulated genes.

## Installation

You can install the current version of **NicheScribe** from GitHub with:

```{r how-to-install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("RabadanLab/NicheScribe")
```

After installation, you will be prompted to download the reference data from the [latest release](https://github.com/RabadanLab/NicheScribe/releases) when you load the package. This is **required** for the package to function properly.  Run the following command:

```{r download-data, eval=FALSE}
download_reference()
```

## Example

We demonstrate NicheScribe using a stromal-enriched bone marrow dataset from [Swann et al. (2026)](https://doi.org/10.1182/blood.2025029513). The input to NicheScribe is a [Seurat v5](https://satijalab.org/seurat/) object with a `counts` matrix in the `RNA` assay.

Note that you may run into out-of-memory errors when working with moderately large scRNA-seq datasets, especially in the UMAP projection step. We suggest increasing the memory limit beforehand:
```{r futures, eval=FALSE}
options(future.globals.maxSize = 100000 * 1024^2)
```

```{r example-swann2026, warning=FALSE, cache=TRUE}
library(NicheScribe)
suppressPackageStartupMessages(library(Seurat))

# Load Swann et al. (2026) dataset from GEO.
options(timeout = 1000)
set3.endo <- ReadMtx(
  cells = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8505738&format=file&file=GSM8505738%5F3EM%5Fbarcodes%2Etsv%2Egz",
  features = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8505738&format=file&file=GSM8505738%5F3EM%5Ffeatures%2Etsv%2Egz",
  mtx = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8505738&format=file&file=GSM8505738%5F3EM%5Fmatrix%2Emtx%2Egz"
)

seu <- CreateSeuratObject(set3.endo, project="Swann2026_endo")

# Normally, the usual data preprocessing and quality control steps are run here but we omit them for brevity.

# Run NicheScribe
seu <- NicheScribe(seu)

# Inspect predictions
table(seu$nonhematopoietic_pred)
table(seu$hash_label_pred)
table(seu$manual_annot_pred)

# Visualize results
NichePlot(seu, layer="R2.1")
NichePlot(seu, layer="R2.2")
InflammationPlot(seu)
```

<!-- 
## Citation

If you use **NicheScribe**, please cite:
-->
